Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-11-17 article_old.tex
Submitted

	In this work, we use HA and NA sequences of A/H3N2 and A/H1N1 influenza from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. 
	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not. 
	To our surprise, we find that the predictability of these trajectories is very limited: The probability that a mutation fixes differs little from its current frequency, as would be expected if fixation happened purely by chance.
	This observation holds for many different categories of mutations, including mutations at epitope positions. 
	This weak predictability is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability. 
	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}. 
	This suggests that although LBI has predictive power, the reason for its success may not be related to it approximating fitness of strains.

@@

The main underlying question asked in this work is the following: given a mutation $X$ in the genome of influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?  \\
There are different ways to make this question more specific. First, one can try to quantitatively predict the frequency of $X$ at future times $f(t)$. In other words, having observed a mutation at frequencies $(f_1, f_2,\ldots,f_n)$ at dates $(t_1,t_2,\ldots,t_n)$, what can we say about its frequency at future dates $(t_{n+1}, t_{n+2},\ldots)$?
A simpler, more qualitative question, is to ask whether $X$ will fix in the population, will disappear, or whether the site will stay polymorphic.

In this work, we use amino-acid sequences of the HA and NA genes of A/H3N2 (since the year 2000) and A/H1N1 (since the year 2001) influenza available in GISAID \citep{shu2017gisaid} (see supplementary materials for an acknowledgment of all data contributors). 
This amounts to $44\,976$ HA and $36\,300$ NA sequences for A/H3N2 and $45\,350$ HA and $40\,412$ NA sequences for A/H1N1, with a minimum of 100 per year. These sequences are binned in non-overlapping intervals of one month. Each single-month time bin and the sequences that it contains represent a (noisy) snapshot of the influenza population at a given date. The number of sequences per time bin varies strongly both with year and according to the season, with earlier time bins containing around 10 sequences while more recent bins contain several hundreds (see figures~\sref{fig:nseq_per_year} and \sref{fig:nseq_per_month} in SM for details).

@@

%Frequency trajectories allow a simple formulation of the above-stated questions. The short term prediction amounts to predicting the curve $f(t)$ starting from a given date $t_0$. On the other hand, the long term prediction amounts to predicting how long a trajectory will remain active, and whether it will fix or disappear when it finishes. 
In the rest of this work, we will focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$. 
These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one. 
If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants. 
In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fixation. 
If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency, should enrich for mutations that are positively selected and on their way to fixation.

@@

To answer this question retrospectively, we use all frequency trajectories extracted from HA and NA sequences that satisfy these conditions for a given $f_0$.  The number of trajectories is limited and the frequency estimates themselves are based on a finite sample and are hence imprecise. 
Therefore, we consider trajectories in an interval $[f_0-\delta f, f_0+\delta f]$ with $\delta f = 0.05$.

For $f_0=0.3$, we find $120$ such trajectories in the case of A/H3N2 influenza, represented on the panel \textbf{B} of figure~\ref{fig:trajectories}, where time is shifted such that $t_0 = 0$. The same analysis is performed for A/H1N1, with the $89$ found trajectories displayed in figure \sref{fig:trajectories_h1n1}.
Some trajectories fall in the frequency bin around $f_0$ while decreasing, even though they crossed that bin at an earlier time. 
This is due to the fact that some trajectories ``skipped" the interval $f_0$ in question on their initial rise due to sparse sampling. 
These trajectories are nevertheless rising in the sense that they start at frequency 0 for $t\rightarrow -\infty$. 
Removing them does not change results significantly.

@@

However, future distributions for $\Delta t >0$ do not seem to behave according to this hypothesis. 
The thick black line in Figure~\ref{fig:trajectories}B shows the average frequency of all trajectories. 
This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$ in the case of A/H3N2, and slightly increasing for A/H1N1.
Hence, the fact that this average rose for $t<0$ gives little information for $t>0$, and is due to the conditions by which these trajectories were selected.
This shows that sweep-like trajectories rising steadily from frequency $0$ to $1$ are not common enough to reflect in the average trajectory.

Consistent with the average, the frequency distribution of the selected trajectories broadens in time without a significant shift of the mean as time passes. 
After 60 days, the distribution is rather symmetrical around the initial $f_0=0.3$ value, suggesting that the knowledge that the trajectories were rising is lost after two months. 
On a timescale of 60 to 120 days, the only possible prediction is that trajectories are likely to be found in a broad interval around the initial frequency $f_0$. 
After one year the distribution becomes almost flat (excluding mutations that have disappeared or fixed), and the initial peak at $f_0$ is not visible anymore. 
The only information remaining from the initial frequency is the fraction that fixed or was lost (see below). 
This behavior is expected in neutral models of evolution \citep{kimura_diffusion_1964} but incompatible with a dynamic dominated by sweeps taking over the population.

While this observation does not rule out that signatures exist that predict future frequency dynamics, past dynamics alone is weakly informative.

@@

\subsection*{Prediction of fixation or loss}
Instead of predicting future frequency, let's consider the more modest goal of predicting the probability that a mutation fixes in the population. 
We first estimate the fraction of frequency trajectories that either fix in the population or are lost, as well as the time it takes for one or the other to happen. 
Panels \textbf{A} and \textbf{B} of figure~\ref{fig:fix_loss_active} shows the fraction of frequency trajectories in HA and NA that either have fixed, were lost or remained active as a function of the time elapsed since they were first seen above 25\% frequency. 
Most mutations are either lost or become fixed after 2-3 years, with very few trajectories remaining active after 5 years.
This time scale of 2-3 years is consistent with the typical coalescence time observed in phylogenetic trees of A/H3N2 influenza \citep{rambaut_genomic_2008,yan_phylodynamic_2019}. 
We also note that the fraction of lost trajectories increases sharply at small times with 40\% of mutations observed above 25\% frequency being lost within one year for A/H3N2, while it takes longer to fix a mutation in the whole population.

@@

We then examined the probability of mutations to fix in the population as a function of the frequency at which they are seen. 
For different values of frequency $f$, we consider all trajectories that started at a null frequency and are seen in the interval $[f - 7.5\%, f + 7.5\%]$ at any given time. 
The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} show $P_{fix}(f)$ as a function of $f$ for NA and HA. 
For both proteins, the probability of fixation of a new mutation at frequency $f$ is very close to $f$ itself, \discuss{that is $P_{fix}(f)\simeq f$.} 
This result is exactly what is expected in a population evolving in the \emph{absence} of or under very weak selection. 
Indeed, in a neutrally evolving and structure-less population of constant size $N$ such as in the Wright-Fisher model~\cite{kimura_diffusion_1964}, the probability that any individual ultimately becomes the ancestor of all the future population is $1/N$. 
A mutation or trait appearing at frequency $f$ is shared by $f\cdot N$ individuals, and the probability for one of them to become the ancestor of all the future population is $f\cdot N/N=f$. 
Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation. 
Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation, while those in A/H1N1 show only small deviations from it. 
In both cases, the probability of fixation seems to be mainly dictated by the current frequency $f$ at which the mutations is observed.

This dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
If strong selection was present, we would expect rising amino acid mutations to fix at a distinctively higher frequency than the one at which they are measured. 
In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.

To investigate further, we try to find categories of mutations that deviate from the diagonal of panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active}. 
We first turn to the \emph{Local Branching Index} (LBI), a quantity calculated for each node in a phylogenetic tree that indicates how dense the branching of the tree is around that node. 
LBI has previously been successfully used as a predictor of the future population of influenza~\cite{neher_predicting_2014}, and was shown to be a proxy for fitness of leaves or ancestral nodes in mathematical models of evolution. 
Here, we define the LBI of a mutation at date $t$ as the average LBI of strains that carry this mutation and that were sampled in the time bin corresponding to $t$. 
Panel A of figure~\ref{fig:pfix_stratified} shows fixation probability for HA mutations with LBI in the top or bottom half of the distribution. 
Both groups have identical probability of fixation, suggesting that LBI carries very little information on the probability of fixation of a mutation.

Next, we focused on previously reported antigenic sites in the A/H3N2 HA protein, referred to as \emph{epitope} positions. 
Mutations at these position might mediate immune escape and are therefore likely under strong selection. 
We used four lists of relevant epitope positions from different sources comprising from 7 to 129 positions in the sequence of the HA1 protein~\cite{Koel976, luksza_predictive_2014, Shih6283, wolf_long_2006}. 
Panel Fig.~\ref{fig:pfix_stratified}\textbf{B} shows fixation probability as a function of frequency for the four lists of epitopes. 
Only mutations at the 7 epitope sites reported in~\cite{Koel976} have higher chances of fixation than expected by chance. 
No clear difference is found for the lists by \citet{luksza_predictive_2014, wolf_long_2006}, while positions from \citet{Shih6283} show lower chances of fixation.
One should also note that many of these positions were determined post-hoc and enriched for positions that experienced rapid substitutions before the publication of the respective studies.

Two ways of categorising mutations, however, suggest some power to predict fixation. 
In panel Fig.~\ref{fig:pfix_stratified}\textbf{C}, we split trajectories into those occurring at binary positions where only two amino acid variants co-circulate and non-binary positions with more than two variants. 
Novel variants at non-binary positions, \emph{i.e.} ones for which competition between three amino acids or more has occurred at least once, have a higher chance of fixation. 
In panel \textbf{D}, we separated mutations that appear more than once or only once in the reconstructed tree (see methods), and found that the former fix more often. 
Panels \textbf{C} and \textbf{D} show that it is possible to gain some information on the chance of fixation of a particular mutation, as was done in panel \textbf{B}. 
However, the predictive power remains small, with the ``top'' curves in panels \textbf{C\&D} being very close to the diagonal.

We conduct the same analysis on A/H1N1 influenza, with results visible in figure \sref{fig:pfix_stratified_h1n1}. 
Results are qualitatively similar to those obtained for A/H3N2, with LBI giving little information and mutations at non-binary positions having a higher chance of fixation. 
Panel \textbf{D} differs between figures~\ref{fig:pfix_stratified} and \sref{fig:pfix_stratified_h1n1}, with convergent evolution giving less information on fixation in the latter case. 
However, this could be due to the shorter time period over which A/H1N1 evolved, resulting in a shorter tree and less possibilities of convergent evolution. 
Indeed, error bars for mutations appearing multiple times in \textbf{D} of figure~\sref{fig:pfix_stratified_h1n1} are relatively large, indicating a lower amount of trajectories.\\

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel3.png}
	\caption{Fixation probability $P_{fix}(f)$ as a function of frequency, for A/H3N2 influenza. Figure \sref{fig:pfix_stratified_h1n1} shows the same analysis for A/H1N1. \textbf{A}: HA mutations with higher or lower LBI values, based on their position with respect to the median LBI value. \textbf{B}: Different lists of epitope positions in the HA protein. The authors and the number of positions is indicated in the legend. \textbf{C}: HA and NA mutations for binary positions, \emph{i.e.} positions for which we never see more than two amino acids in the same time bin. \textbf{D}: HA and NA mutations that appear once or more than once in the tree for a given time bin.}
	\label{fig:pfix_stratified}
\end{figure*}

@@

The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the notion that seasonal influenza virus evolution is driven by rapid directed positive selection.
One possible explanation for the weakly predictable behaviour of mutations (beyond current frequency) might be tight genetic linkage inside each protein and strong competition between different adaptive mutations \citep{neher_genetic_2011}. 
We design a simple model of population evolution based on the \texttt{ffpopsim} simulation software to test this hypothesis~\cite{10.1093/bioinformatics/bts633}. 
The model represents a population of binary genomes of length $L=200$ evolving in a fitness landscape that changes through time.

@@

Having simulated populations in these two fitness landscapes, we perform the same analysis of frequency trajectories as for the real influenza data. 
Figure~\sref{fig:simulations} of the SM shows the $P_{fix}(f)$ as a function of $f$ for the two models and for different values of the inverse rate of change $\Delta t$ of the fitness landscape. 
For all models, this curve deviates significantly from the diagonal. 
This is most evident for the case of a simple additive fitness landscape that changes rarely $\Delta t=1000$: rising mutations almost always fix in the population, with $P_{fix}(f)\simeq 1$ for any $f$ larger than a few percent. 
This is corroborated by visual inspection of the trajectories, which shows that evolution in this regime is driven by regular selective sweeps that take a typical time of $\sim 400$ generations. 
In other regimes, with smaller $\Delta t$ or with strong epistatic competition, $P_{fix}(f)$ is reduced and closer to the diagonal. 
However, it takes an extremely fast changing fitness landscape to push $P_{fix}$ close to the diagonal: with $\Delta t=10$, that is about 40 changes to the fitness landscape in the time it would take a selective sweep to go from 0\% to fixation, $P_{fix}(f)$ differs from $f$ in a way that is comparable to what is observed in A/H1N1 influenza.

These models are not meant to be accurate models of influenza viruses evolution. 
But figure~\sref{fig:simulations} does show is that the patterns observed in influenza virus evolution are only reproduced by models of adapting populations when pushing clonal competition to extreme values.
We conclude that the pattern in figure~\ref{fig:fix_loss_active} may not a straightforward manifestation of genetic linkage and clonal interference, but that some more intricate interplay of epidemiology, seasonality, human immunity and chance gives rise to the weakly predictable yet strongly selected evolutionary dynamics of IAVs.

\subsection*{Why do predictions work?}
The statistics of frequency trajectories seems to be in conflict with the notion that influenza evolution is predictable.
Likewise, the LBI, a quantity that correlates with fitness in mathematical models and is used to predict future influenza populations \citep{neher_predicting_2014}, does not seem to contain any information on whether a specific mutation is going to fix or not, see figure~\ref{fig:pfix_stratified}. 
To resolve this conundrum, we first note that criterion by which predictive power for influenza was measured in \citep{neher_predicting_2014} was the distance between the strain with the highest LBI and the future population, not the ability of the LBI to predict dynamics.
The distance was compared to the average distance between the present and future population, as well as the post-hoc optimal representative and the future.

@@

To assess the ability of the LBI to pick a close representative of the future, we compute the LBI of each node of one time bin in the tree using only the leaves that belong to that time bin. 
The top panel in figure~\ref{fig:LBI_consensus} shows the hamming distance of the strain with the highest LBI to future populations at different $\Delta t$ along with the same distance for a randomly chosen strain. 
The figure shows the distance averaged over all possible values of $t$ for $\Delta t$ between $0$ and $32$ months, giving us an average efficiency of a predictor over 16 years of influenza evolution.

The strain with the highest LBI is consistently closer to the future than the average strain by about 1-2 amino acids, while the overall distance increases linearly due to the continuous evolution of the population.
We hence reproduce previous results showing that the LBI picks closer than average representatives \citep{neher_predicting_2014}.
To investigate whether this apparent success is due to the ability of the LBI to predict fitness or not, we explored a different predictor: the amino acid consensus sequence of the present population (see Methods for a definition of the consensus sequence). 
The choice is motivated by the fact that it can be shown to be the best possible long term predictor for a neutrally evolving population in terms of Hamming distance (see SM section \ref{sub:consensus_sequence_as_a_predictor}). 
Figure~\ref{fig:LBI_consensus} shows that the consensus sequence is in fact a equally good or even slightly better representative of the future than the sequence with highest LBI 
(note that the consensus sequence does \emph{not} necessarily exist in the population).

@@

The premises of predicting the trajectory of a mutation are (i) significant fitness difference between genomes carrying different variants at the site and (ii) a selection pressure that changes slowly over time. 
Under such conditions, it is expected that frequency trajectories will show a persistent behavior which would make them predictable for some time. 
However, we could find only limited evidence for such persistent behavior in the past 19 years of IAV evolution. 
This lead us to conclude that (i) influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) previous methods to predict influenza evolution work primarily because they pick strains that represent the future well, not because they predict future dynamics.

>>>>

	In this work, we use A/H3N2 HA and NA sequences from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. 
	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not. 
	To our surprise, we find that the predictability of these trajectories is very limited: The probability that a mutation fixes differs little from its current frequency as would be expected if fixation happened purely by chance.
	This observation holds for many different categories of mutations, including mutations at epitope positions. 
	This quasi-neutrality is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability. 
	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}. 
	This suggests that although LBI has predictive power, the reason for its success may not be related to it approximating fitness of strains.

@@

The main underlying question asked in this work is the following: given a mutation $X$ in the genome of A/H3N2 influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?  \\
There are different ways to make this question more specific. First, one can try to quantitatively predict the frequency of $X$ at future times $f(t)$. In other words, having observed a mutation at frequencies $(f_1, f_2,\ldots,f_n)$ at dates $(t_1,t_2,\ldots,t_n)$, what can we say about its frequency at future dates $(t_{n+1}, t_{n+2},\ldots)$?
A simpler, more qualitative question, is to ask whether $X$ will fix in the population, will disappear, or whether the site will stay polymorphic.

In this work, we use amino-acid sequences of the HA and NA genes of A/H3N2 influenza since the year 2000 available in GISAID \citep{shu2017gisaid} (see supplementary materials for an acknowledgment of all data contributors). 
This amounts to $44\,976$ HA and $36\,300$ NA sequences, with a minimum of 100 per year. These sequences are binned in non-overlapping intervals of one month. Each single-month time bin and the sequences that it contains represent a (noisy) snapshot of the A/H3N2 population at a given date. The number of sequences per time bin varies strongly both with year and according to the season, with earlier time bins containing around 10 sequences while more recent bins contain several hundreds (see figures~\sref{fig:nseq_per_year} and \sref{fig:nseq_per_month} in SM for details).

@@

%Frequency trajectories allow a simple formulation of the above-stated questions. The short term prediction amounts to predicting the curve $f(t)$ starting from a given date $t_0$. On the other hand, the long term prediction amounts to predicting how long a trajectory will remain active, and whether it will fix or disappear when it finishes. 
In the rest of this work, we will focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$. 
These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one. 
If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants. 
In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e. the rapid rise and fixation. 
If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency, should enrich for mutations that are positively selected and on their way to fixation.

@@

To answer this question retrospectively, we use all frequency trajectories extracted from A/H3N2 HA sequences that satisfy these conditions for a given $f_0$.  The number of trajectories is limited and the frequency estimates themselves are based on a finite sample and are hence imprecise. 
Therefore, we consider trajectories in an interval $[f_0-\delta f, f_0+\delta f]$ with $\delta f = 0.05$.

For $f_0=0.3$, we find 69 such trajectories, represented on the panel \textbf{B} of figure~\ref{fig:trajectories}, where time is shifted such that $t_0 = 0$. 
Some trajectories fall in the frequency bin around $f_0$ while decreasing, even though they crossed that bin at an earlier time. 
This is due to the fact that some trajectories ``skipped" the interval $f_0$ in question on their initial rise due to sparse sampling. 
These trajectories are nevertheless rising in the sense that they start at frequency 0 for $t\rightarrow -\infty$. 
Removing them does not change results significantly.

@@

However, future distributions for $\Delta t >0$ do not seem to behave according to this hypothesis. 
The thick black line in Figure~\ref{fig:trajectories}B shows the average frequency of all trajectories. 
This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$.
The fact that this average rose $t<0$ has hence no information for $t>0$ and is due to the conditions by which these trajectories were selected.

Consistent with the average, the frequency distribution of the selected trajectories broadens in time without a significant shift of the mean as time passes. 
After 60 days, the distribution is rather symmetrical around the initial $f_0=0.3$ value, suggesting that the knowledge that the trajectories were rising is lost after two months. 
On a timescale of 60 to 120 days, the only possible prediction is that trajectories are likely to be found in a broad interval around the initial frequency $f_0$. 
After one year the distribution becomes almost flat (excluding mutations that have disappeared or fixed), and the initial peak at $f_0$ is not visible anymore. 
The only information remaining from the initial frequency is the fraction that fixed or was lost (see below). 
This behavior is expected in neutral models of evolution \citep{kimura_diffusion_1964} but incompatible with the notion of classic sweeps which take over the population.

While this observation does not rule out that signatures exist that predict future frequency dynamics, past dynamics alone is uninformative.

@@

\subsection*{Prediction of fixation or loss}
Instead of predicting future frequency, let's consider the more modest goal of predicting the probability that a mutation fixes in the population. 
We first estimate the fraction of frequency trajectories that either fix in the population or are lost, as well as the time it takes for one or the other to happen. 
Panels \textbf{A} and \textbf{B} of figure~\ref{fig:fix_loss_active} shows the fraction of frequency trajectories in HA and NA that either have fixed, were lost or remained active as a function of the time elapsed since they were first seen above 25\% frequency. 
Most mutations are either lost or become fixed after 2-3 years, with very few trajectories remaining active after 5 years.
This time scale of 2-3 years is consistent with the typical coalescence time observed in phylogenetic trees of A/H3N2 influenza \citep{rambaut_genomic_2008,yan_phylodynamic_2019}. 
We also note that the fraction of lost trajectories increases sharply at small times, indicating that 40\% of mutations observed above 25\% frequency are lost within one year, while it takes longer to fix a mutation in the whole population.

@@

We then examined the probability of mutations to fix in the population as a function of the frequency at which they are seen. 
For different values of frequency $f$, we consider all trajectories that started at a null frequency and are seen in the interval $[f - 7.5\%, f + 7.5\%]$ at any given time. 
The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} shows $P_{fix}(f)$ as a function of $f$ for NA and HA and for influenza A/H3N2 and A/H1N1. 
\discuss{For both these proteins, the probability of fixation of a new mutation at frequency $f$ is very close to $f$ itself, that is $P_{fix}(f)\simeq f$.} 
This result is exactly what is expected in a population evolving in the \emph{absence} of selection. 
Indeed, in a neutrally evolving and structure-less population of constant size $N$ such as in the Wright-Fisher model~\cite{kimura_diffusion_1964}, the probability that any individual ultimately becomes the ancestor of all the future population is $1/N$. 
A mutation or trait appearing at frequency $f$ is shared by $f\cdot N$ individuals, and the probability for one of them to become the ancestor of all the future population is $f\cdot N/N=f$. 
Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation. 
Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation. \discuss{Add smth for A/H1N1}

This quasi-neutral dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
If strong selection was present, we would expect rising amino acid mutations to fix at a higher frequency than the one at which they are measured. 
In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.

To investigate further, we try to find categories of mutations that deviate from the neutral expectation of panel \textbf{C} of figure~\ref{fig:fix_loss_active}. 
We first turn to the \emph{Local Branching Index} (LBI), a quantity calculated for each node in a phylogenetic tree that indicates how dense the branching of the tree is around that node. 
LBI has previously been successfully used as a predictor of the future population of influenza~\cite{neher_predicting_2014}, and was shown to be a proxy for fitness of leaves or ancestral nodes in mathematical models of evolution. 
Here, we define the LBI of a mutation at date $t$ as the average LBI of strains that carry this mutation and that were sampled in the time bin corresponding to $t$. 
Panel A of figure~\ref{fig:pfix_stratified} shows fixation probability for HA mutations with LBI in the top or bottom half of the distribution. 
Both groups are again consistent with neutral evolution, suggesting that LBI carries very little information on the probability of fixation of a mutation.

Next, we focused on previously reported antigenic sites in the HA protein, referred to as \emph{epitope} positions. 
Mutations at these position might mediate immune escape and are therefore likely under strong selection. 
We used four lists of relevant epitope positions from different sources comprising from 7 to 129 positions in the sequence of the HA1 protein~\cite{Koel976, luksza_predictive_2014, Shih6283, wolf_long_2006}. 
Panel Fig.~\ref{fig:pfix_stratified}\textbf{B} shows fixation probability as a function of frequency for the four lists of epitopes. 
Only mutations at the 7 epitope sites reported in~\cite{Koel976} have higher chances of fixation than expected by chance. 
No clear difference is found for the lists by \citet{luksza_predictive_2014, wolf_long_2006}, while positions from \citet{Shih6283} show lower chances of fixation.
One should also note that many of these positions were determined post-hoc and enriched for positions that experienced rapid substitutions before the publication of the respective studies.

Two ways of categorising mutations, however, suggest some power to predict fixation. 
In panel Fig.~\ref{fig:pfix_stratified}\textbf{C}, we split trajectories into binary trajectories where only two amino acid variants co-circulate and non-binary ones with more than two variants. 
Novel variants at non-binary positions, \emph{i.e.} ones for which competition between three amino acids or more has occurred at least once, have a higher chance of fixation. 
In panel \textbf{D}, we separated mutations that appear more than once or only once in the reconstructed tree (see methods), and found that the former fix more often. 
Panels \textbf{C} and \textbf{D} show that it is possible to gain some information on the chance of fixation of a particular mutation, as was done in panel \textbf{B}. 
However, the predictive power remains small, with the ``top'' curves in panels \textbf{C\&D} being very close to the diagonal.

We conduct the same analysis on A/H1N1 influenza, with results visible in figure \sref{fig:pfix_stratified_h1n1}. 
Results are qualitatively similar to those obtained for A/H3N2, with LBI giving little information and mutations at non-binary positions having a higher chance of fixation. 
Panel \textbf{D} differs between figures~\ref{fig:pfix_stratified} and \sref{fig:pfix_stratified_h1n1}, with convergent evolution giving less information on fixation in the latter case. 
However, this could be due to the shorter time period over which A/H1N1 evolved, resulting in a shorter tree and less possibilities of convergent evolution. 
Indeed, error bars for mutations appearing multiple times in \textbf{D} of figure~\sref{fig:pfix_stratified_h1n1} are relatively large, indicating a lower amount of trajectories.

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel3.png}
	\caption{Fixation probability $P_{fix}(f)$ as a function of frequency. \textbf{A}: Mutation with higher or lower LBI values, based on their position with respect to the median LBI value. \textbf{B}: Different lists of epitope positions in the HA protein. The authors and the number of positions is indicated in the legend. \textbf{C}: Mutations for binary positions, \emph{i.e.} positions for which we never see more than two amino acids in the same time bin. \textbf{D}: Mutations that appear once or more than once in the tree for a given time bin.}
	\label{fig:pfix_stratified}
\end{figure*}

@@

The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the notion that seasonal influenza virus evolution is driven by rapid directed positive selection.
One possible explanation for the quasi-neutral behaviour of mutations might be tight genetic linkage inside each protein and strong competition between different adaptive mutations \citep{neher_genetic_2011}. 
We design a simple model of population evolution based on the \texttt{ffpopsim} simulation software to test this hypothesis~\cite{10.1093/bioinformatics/bts633}. 
The model represents a population of binary genomes of length $L=200$ evolving in a fitness landscape that changes through time.

@@

Having simulated populations in these two fitness landscapes, we perform the same analysis of frequency trajectories as for the real influenza data. 
Figure~\sref{fig:simulations} of the SM shows the $P_{fix}(f)$ as a function of $f$ for the two models and for different values of the inverse rate of change $\Delta t$ of the fitness landscape. 
For all models, this curve deviates significantly from the diagonal. 
This is most evident for the case of a simple additive fitness landscape that changes rarely $\Delta t=1000$: rising mutations almost always fix in the population, with $P_{fix}(f)\simeq 1$ for any $f$ larger than a few percent. 
This is corroborated by visual inspection of the trajectories, which shows that evolution in this regime is driven by regular selective sweeps that take a typical time of $\sim 400$ generations. 
In other regimes, with smaller $\Delta t$ or with strong epistatic competition, $P_{fix}(f)$ is reduced and closer to the diagonal. 
However, even in an extremely fast changing fitness landscape with $\Delta t=10$, that is about 40 changes to the fitness landscape in the time it would take a selective sweep to go from 0\% to fixation, $P_{fix}(f)$ still significantly differs from $f$.

These models are not meant to be accurate models of influenza viruses evolution. 
\discuss{But figure~\sref{fig:simulations} does show is that the patterns observed in influenza virus evolution are not readily reproduced by in models of adapting populations, even if they include strong clonal competition and epistasis.
We conclude that the pattern in figure~\ref{fig:fix_loss_active} is not a straightforward manifestation of genetic linkage and clonal interference, but that some more intricate interplay of epidemiology, seasonality, human immunity and chance gives rise to the quasi-neutral yet strongly selected evolutionary dynamics of IAVs.}

\subsection*{Why do predictions work?}
The quasi-neutral statistics of frequency trajectories seems to be in conflict with the notion that influenza evolution is predictable.
Likewise, the LBI, a quantity that correlates with fitness in mathematical models and is used to predict future influenza populations \citep{neher_predicting_2014}, does not seem to contain any information on whether a specific mutation is going to fix or not, see figure~\ref{fig:pfix_stratified}. 
To resolve this conundrum, we first note that criterion by which predictive power for influenza was measured in \citep{neher_predicting_2014} was the distance between the strain with the highest LBI and the future population, not the ability of the LBI to predict dynamics.
The distance was compared to the average distance between the present and future population, as well as the post-hoc optimal representative and the future.

@@

To assess the ability of the LBI to pick a close representative of the future, we compute the LBI of each node of one time bin in the tree using only the leaves that belong to that time bin. 
The top panel in figure~\ref{fig:LBI_consensus} shows the hamming distance of the strain with the highest LBI to future populations at different $\Delta t$ along with the same distance for randomly chosen strain. 
The figure shows the distance averaged over all possible values of $t$ for $\Delta t$ between $0$ and $32$ months, giving us an average efficiency of a predictor over 16 years of influenza evolution.

The strain with the highest LBI is consistently closer to the future than the average strain by about 1-2 amino acids, while the overall distance increases linearly due to the continuous evolution of the population.
We hence reproduce previous results showing that the LBI picks closer than average representatives \citep{neher_predicting_2014}.
To investigate whether this apparent success is due to the ability of the LBI to predict fitness or not, we explored a different measure: the amino acid consensus sequence of the present population (see Methods for a definition of the consensus sequence). 
The choice is motivated by the fact that it can be shown to be the best possible long term predictor for a neutrally evolving population in terms of Hamming distance (see SM section \ref{sub:consensus_sequence_as_a_predictor}). 
Figure~\ref{fig:LBI_consensus} shows that the consensus sequence is in fact a equally good or even slightly better representative of the future than the sequence with highest LBI 
(note that this sequence does \emph{not} necessarily exist in the population).

@@

The premises of predicting the trajectory of a mutation are (i) significant fitness difference between genomes carrying different variants at the site and (ii) a selection pressure that changes slowly over time. 
Under such conditions, it is expected that frequency trajectories will show a persistent behavior which would make them predictable for some time. 
However, we could find only limited evidence for such persistent behavior in the past 19 years of A/H3N2 evolution. \pierre{But we do find evidence for A/H1N1, not of a large magnitude.}
This lead us to conclude that (i) A/H3N2 influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) previous methods to predict influenza evolution work primarily because they pick strains that represent the future well, not because they predict future dynamics.
